A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2022-06-20 paper/paper.tex
More updates on paper

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}

@@

%%
%% end of the preamble, start of the body of the document source.
\begin{document}
%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Towards a Sound and Complete Library for Modal Logics Reasoning in Coq}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Modal logics, soundness, completeness, Coq proof assistant}

@@

Our society depends heavily on computing technology: our cars, airplanes, medical devices
and banking applications all rely on software artifacts. In this scenario,
\emph{formal methods} play a crucial role to ensure that a program follows its
specification, thus avoiding failures which can cause castrophic outcomes. Interactive
proof assistants, like Coq~\cite{manual_coq} and Isabelle/HOL~\cite{manual_isabelle}, can
be used to verify software components and have been applied with success in
non-trivial projects like a certified compiler for C programming
language~\cite{Leroy09} and the kernel of seL4 operating system~\cite{Klein10}.

Usually, complex software systems are formed by various inter-communicating
components. A key aspect of the design of such programs is how to ensure that
a failure of a single component does not compromise the whole system.
The main difficulty is such formalization tasks is how to guarantee correct
error detection and that its corresponding fix will ensure that the software
still continue to work under abnormal conditions. One way to express such
requirements is using \emph{modal logics}, which can specify communication
patterns and knowledge in software agents~\cite{Huth_Ryan}.

While modal logics allow the modelling of such requirements, proving properties
about them may be a hard task without proper tool support. There are some
dedicated provers for modal logics~\cite{Gleissner17,Mora11}, but we believe
that using mature tools, like a proof assistant, are a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newbies to join the game of formal verification and 3) The custom automation
facilities supported by current proof assistants allows that trivial facts can
be proved using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

@@

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive systems as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL
    and prove its soundness with respect to Kripke semantics. We also
    discuss the current status of the formalization of the completeness result.
  \item We show a proof of the L\"ob's theorem using the Hilbert deductive system
    developed in our library.
\end{itemize}

The rest of this work is organized as follows: Section~\ref{sec:background} provides
brief introduction on modal logics and the Coq proof assistant. Details about
the implementation of our Coq library are presented in
Section~\ref{sec:implementation}. A case study in proving the L\"ob's theorem is
the subject of Section~\ref{sec:case}. Related work is discussed
on Section~\ref{sec:related} and Section~\ref{sec:conclusion} draws some conclusions and
provide some pointers for future works.

@@

In this work, we consider Hilbert deductive systems for modal logics.
Essentially, the considered system adds axioms for the modalities
to the set of axioms for classical propositional logic.
As examples of modalities axioms we mention \textbf{K}, which express that $\Box$ distributes
over implication, and possibility which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    \textbf{K} & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    \textbf{Possibility} & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_2 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need the traditional \emph{modus ponens} rule
and a rule that allows to conclude formulas with necessity.

@@

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by a
a type \mintinline{coq}{W : Set} to represent the set of worlds and the
accessibility relation is denoted by a type \mintinline{coq}{R : W -> W -> Prop}.
\begin{minted}{coq}
Record Frame : Type := {
  W : Set;
  R : W -> W -> Prop;
}.
\end{minted}
A model is denoted by another record formed by a frame and a labelling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model : Type := {
  F : Frame; 
  v : nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labelling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}. The encoding of the Kripke semantics is done by a Coq function
that which implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labelling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ are represented by a quantifying over worlds
using the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model)
                           (w: W (F M))
                           (phi: form): Prop :=
  match phi with
  | Lit     x            => v M x w 
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                   formula_valuation M w' psi
  (** some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimics the standard way of writing them.

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of on four
possibilities in a proof: 1) the use of a hypothesis; 2) use of an axiom;
3) use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used by several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discuss the rule for axioms, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom : Set :=
  | ax1   : modalFormula -> modalFormula -> axiom
  | ax2   : modalFormula -> modalFormula ->
                            modalFormula -> axiom
  (** some code omitted *)
  | axK   : modalFormula -> modalFormula -> axiom
  | axPos : modalFormula -> modalFormula -> axiom
  | axT   : modalFormula -> axiom
  | axB   : modalFormula -> axiom
  | axK4  : modalFormula -> axiom
  | axD   : modalFormula -> axiom
  | axK5  : modalFormula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and build its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (** some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

The \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate} to
create a formula instance from using a given system, represented as a predicate
\mintinline{coq}{A : axiom -> Prop}, which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e. it is not empty. Next, we
detail the representation of different systems of modal logic in our library.

@@

While we have completely proved the soundness theorem, the current version of
our library does not have a completeness theorem.

@@

Using the definition of fixed points, we can state L\"ob's theorem as follows:
\begin{minted}{coq}
Theorem Lob:
  forall P, fixed_point K4 nil ->
  (K4; nil |-- [! []P -> P !]) ->
  (K4; nil |-- [! P !]).
\end{minted}

The proof starts by assuming an arbitrary formula $P$, the existence of modal fixed
points and that $\Box P \to P$ is provable in K4. These deduction steps
are performed by the following tactic:
\begin{minted}{coq}
intros P FP H1.
\end{minted}

Next, from the existence of modal fixed points for every formula,
we know that there is a formula $\Psi$ such that $\Psi \leftrightarrow (\Box
\Psi \to P)$. The following tactic performs this step of reasoning.
\begin{minted}{coq}
destruct FP with (fun A => [! A -> P !]) as (psi, H2).
\end{minted}
Now, we are able to deduce that $\Psi \to (\Box \Psi\to P)$ from the previous
step result as follows:
\begin{minted}{coq}
assert (K4; nil |-- [! psi -> []psi -> P !]) as H3.
apply modal_ax5 in H2; try repeat constructor ;
   try eassumption.
\end{minted}
The \mintinline{coq}{assert} tactic allows the definition of
auxiliary proof goals. In the previous code piece, we specified
$\vdash \Psi \to (\Box \Psi \to P)$ as a new goal which
generates a new proof hypothesis named \mintinline{coq}{H3}. The
deduction of this fact is immediate from K4 axioms.

@@

Now, using the axiom \textbf{K} we can deduce $\vdash \Box \Psi \to
\Box (\Box \Psi \to P)$. Again, we deduce this fact using the
\mintinline{coq}{assert} tactic.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> []([]psi -> P) !]) as H5.
apply modal_axK in H4; try repeat constructor ;
  try eassumption.
\end{minted}

The next proof step involves showing that
$\vdash \Box (\Box \Psi \to P)\to (\Box \Box \Psi \to \Box P)$ holds
by axiom \textbf{K}.
\begin{minted}{coq}
assert (K4; nil |-- [! []([]psi -> P) ->
                       [][]psi -> []P !]) as H6.
eapply Ax with (a := axK ?[X] ?[Y]);
  try repeat constructor.
\end{minted}

Using $\vdash \Box (\Box \Psi \to P)\to (\Box \Box \Psi \to \Box P)$ and
$\vdash \Box \Psi \to\Box (\Box \Psi \to P)$ we can deduce
$\vdash \Box \Psi \to \Box \Box \Psi \to \Box P$ using composition of
the implication.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> [][]psi -> []P !]) as H7.
eapply modal_compose; try repeat constructor ;
   try eassumption.
\end{minted}

Now, we can show that $\vdash\Box\Psi\to\Box\Box\Psi$ using
the axioms of K4. The next snippet of tactics perform this
reasoning step.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> [][]psi !]) as H8.
apply modal_axK4; try constructor 2.
\end{minted}

From $\vdash \Box\Psi \to \Box P$ and $\vdash \Box P \to P$, we can
conclude $\vdash \Box \Psi \to P$, by composition of implication as
follows.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> P !]) as H10.
eapply modal_compose; try repeat constructor ;
   try eassumption.
\end{minted}

Next, we can conclude $\vdash \Box (\Psi \to P) \to \Psi$ from
$\Psi \leftrightarrow (\Box \Psi \to P)$ and the axioms of K4 using
the following tactics:
\begin{minted}{coq}
assert (K4; nil |-- [! ([]psi -> P) -> psi !]) as H11.
apply modal_ax6 in H2; try repeat constructor ;
 try eassumption.
\end{minted}

From $\vdash (\Box\Psi \to P) \to \Psi$ and $\vdash \Box \Psi \to P$
we can conclude $\vdash \Psi$ by modus ponens.
\begin{minted}{coq}
assert (K4; nil |-- psi) as H12.
eapply Mp ; try eassumption.
\end{minted}

By necessitation rule, we can deduce $\vdash \Box \Psi$
\begin{minted}{coq}
assert (K4; nil |-- [! []psi !]) as H13.
apply Nec ; try eassumption.
\end{minted}

Finally, the desired conclusion $\vdash P$ follows from
using the rule modus ponens over $\vdash \Box \Psi \to P$
and $\vdash \Box \Psi$, as presented by the next tactics
snippet.
\begin{minted}{coq}
eapply Mp ; try eassumption.
\end{minted}

This concludes the formalization of the L\"ob's theorem using
the developed library.

@@

A model for recursive, impredicatively quantified types with mutable references
was studied by Appel et. al.~\cite{Appel07}. His interest in formalizing such
advanced typing concepts was in the context of typed assembly languages.
Authors stabilished a semantic type soundness proof by defining a Kripke-style 
semantics of the G\"odel-L\"ob logic of provability. All their definitions were
mechanized through a shallow embedding in Coq.

The embedding of modal linear logic in Coq was the subject of
Sadrzadeh's work~\cite{Soton03}. Like our work, Sadrzadeh's uses a
deep-embedding approach and assumes that all modalities are reflexive
and transitive. Using her embedding, Sadrzadeh shows some proofs of classical
logic puzzles.

@@

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,anonymous]{acmart}

@@

%%
%% end of the preamble, start of the body of the document source.
\begin{document}
%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Modal logics, soundness, Coq proof assistant}

@@

The start of modal logics can be traced back to Lewis and his definition of 
logical systems with strict implication instead of material implication \cite{Lewis1918}.
Based in this work~\cite{Lewis1932}, independently of classical propositional logic,
defined the systems $\mathsf{S1}$--$\mathsf{S5}$, where $\mathsf{S4}$ and $\mathsf{S5}$
are well known normal modal systems. It was, in fact, G{\"o}del that had the idea 
to build a modal logic system as an extension of the classical propositional calculus, 
when constructing a system similar to $\mathsf{S4}$ in \cite{Godel1933}.

Despite that the original motivation of modal systems was to deal with strict implication, 
for many years modal logic was seen as ``the logic of necessity and possibility''. 
Only after the introduction of Kripke semantics they were seen also as ``simple yet
expressive languages for talking about relational structures''~\cite{Blackburn-ModalLogics}. 
A great multitude of modal systems, with various kinds of modal operators, each of them with 
different meanings or properties, have then been constructed in order to provide effective
formalisms for talking about time, space, knowledge, beliefs, actions, 
obligations, among others.

The interpretation of modal logic simply as ``the logic of necessity'' restricts 
not only its applications but also the range of modal systems. In fact, modal logic 
can also be seen as ``the study of \emph{modalities} -- logical operations that qualify 
assertions about the truth of statements''~\cite{Goldblatt-MathofModality}. In this sense, 
the necessitation rule (from $\varphi$ infers $\Box\varphi$) can be relaxed. This brings 
a number of modal systems known as non-normal, whose semantics was defined only in the 
1970's by Montague~\cite{Montague1970} and Scott~\cite{Scott_AdviceModal1970}.

Such wide range of highly expressive logical languages is the ideal tool to 
formalize the many complexities of modern advanced software systems, 
such as those found in cars, airplanes, medical devices and banking applications.
In combining said modal logics with interactive proof assistants,
such as Coq~\cite{manual_coq} and Isabelle/HOL~\cite{manual_isabelle}, both of
which have been applied with success in non-trivial projects like a certified compiler 
for C programming language~\cite{Leroy09} and the kernel of seL4 operating
system~\cite{Klein10}, it is possible to formalize and verify highly complex computer
systems.

Although there are some dedicated provers for modal logics~\cite{Gleissner17,Mora11},
we believe that using mature tools, like a proof assistant, are a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newbies to join the game of formal verification and 3) The custom automation
facilities supported by current proof assistants allows that trivial facts can
be proved using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

@@

% Our society depends heavily on computing technology: our cars, airplanes, medical devices
% and banking applications all rely on software artifacts. In this scenario,
% \emph{formal methods} play a crucial role to ensure that a program follows its
% specification, thus avoiding failures which can cause castrophic outcomes. Interactive
% proof assistants, like Coq~\cite{manual_coq} and Isabelle/HOL~\cite{manual_isabelle}, can
% be used to verify software components and have been applied with success in
% non-trivial projects like a certified compiler for C programming
% language~\cite{Leroy09} and the kernel of seL4 operating system~\cite{Klein10}.

% Usually, complex software systems are formed by various inter-communicating
% components. A key aspect of the design of such programs is how to ensure that
% a failure of a single component does not compromise the whole system.
% The main difficulty is such formalization tasks is how to guarantee correct
% error detection and that its corresponding fix will ensure that the software
% still continue to work under abnormal conditions. One way to express such
% requirements is using \emph{modal logics}, which can specify communication
% patterns and knowledge in software agents~\cite{Huth_Ryan}.

% While modal logics allow the modelling of such requirements, proving properties
% about them may be a hard task without proper tool support. There are some
% dedicated provers for modal logics~\cite{Gleissner17,Mora11}, but we believe
% that using mature tools, like a proof assistant, is a better option for
% verifying facts using modal logics. We justify this choice by: 1) the recent
% successful histories on verification of software and
% mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
% community of developers using such tools thanks to 
% textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
% newbies to join the game of formal verification and 3) The custom automation
% facilities supported by current proof assistants allows that trivial facts can
% be proved using decision procedures, relieving the user from writing proof
% scripts for formulas that can be checked automatically.

% In this context, the present work aims to contribute by constructing a library
% that formalize a family of normal modal logics and its related properties in Coq
% by parametrizing the proofs over a set of possible axioms. 
% In using such a library, a developer could use all Coq's resources to prove
% modal logics facts.

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive systems as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL
    and prove its soundness with respect to Kripke semantics.
    We also formalize several frame correspondence results.
  \item We show a proof of the L\"ob's theorem using the Hilbert deductive system
    developed in our library.
\end{itemize}

The rest of this work is organized as follows: Section~\ref{sec:background} provides
brief introduction on modal logics and the Coq proof assistant. Details about
the implementation of our Coq library are presented in
Section~\ref{sec:implementation}. A case study in proving the L\"ob's theorem is
the subject of Section~\ref{sec:case}. Related works are discussed
on Section~\ref{sec:related} and Section~\ref{sec:conclusion} draws some conclusions and
provide some pointers for future works.

@@

In this work, we consider Hilbert deductive systems for modal logics.
Essentially, the considered system adds axioms for the modalities
to the set of axioms for classical propositional logic.
As examples of modalities axioms we mention \textbf{K}, which express that $\Box$ distributes
over implication, and possibility which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    \textbf{K} & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    \textbf{Possibility} & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_2 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need \emph{modus ponens} and necessitation rules.

@@

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by a
a type \mintinline{coq}{W : Set} to represent the set of worlds and the
accessibility relation is denoted by a type \mintinline{coq}{R : W -> W -> Prop}.
\begin{minted}{coq}
Record Frame : Type := {
  W : Set;
  R : W -> W -> Prop;
}.
\end{minted}
A model is denoted by another record formed by a frame and a labelling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model : Type := {
  F : Frame; 
  v : nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labelling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}. The encoding of the Kripke semantics is done by a Coq function
that implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labelling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ are represented by a quantifying over worlds
using the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model)
                           (w: W (F M))
                           (phi: form): Prop :=
  match phi with
  | Lit     x            => v M x w 
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                   formula_valuation M w' psi
  (** some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimics the standard way of writing them.

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of on four
possibilities in a proof: 1) the use of a hypothesis; 2) use of an axiom;
3) use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used by several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discuss the rule for axioms, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom : Set :=
  | ax1   : form -> form -> axiom
  | ax2   : form -> form -> form  -> axiom
  (** some code omitted *)
  | axK   : form -> form -> axiom
  | axPos : form -> form -> axiom
  | axT   : form -> axiom
  | axB   : form -> axiom
  | axK4  : form -> axiom
  | axD   : form -> axiom
  | axK5  : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and build its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (** some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

The \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate}
function to create a formula instance from using a given system,
represented as a predicate
\mintinline{coq}{A : axiom -> Prop}, which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e. it is not empty. Next, we
detail the representation of different systems of modal logic in our library.

@@

\paragraph{Frame Correspondence} 
As already mentioned, there are some modal formula schemas that correspond
to properties of the accessibility relation of a modal system, if we take them
as an axiom in the system.  While the proof that a certain property in a frame
implies the correspondent axiom is usually  direct, the inverse often needs
more effort. On the Coq proofs developed for this work,  the former case is
mostly unfoldings, simplifications and applications of premises
while the latter, appart from the necessity and burden of a proof by contraposition, 
requires the construction of a witness valuation function. 
Nevertheless, all Coq's proofs are fairly similar to the respective
``pen and paper'' proofs that  can be found in the literature. 
Our work proves the correspondence for the  following properties in
accessibility relations: reflexivity, transitivity, symmetry,  euclidean,
seriality, functionality, density, convergency, and converse well-foundness.

@@

\noindent Using the definition of fixed points, we can state L\"ob's theorem as follows:

\begin{minted}{coq}
  Theorem Lob: forall P, fixed_point K4 nil ->
               (K4; nil |-- [! []P -> P !]) ->
               (K4; nil |-- [! P !]).
\end{minted}

\noindent The proof starts by assuming an arbitrary formula $P$, the existence of modal fixed
points and that $\Box P \to P$ is provable in K4. These deduction steps
are performed by the following tactic:

\begin{minted}{coq}
  intros P FP H1.
\end{minted}

\noindent Next, from the existence of modal fixed points for every formula,
we know that there is a formula $\Psi$ such that $\Psi \leftrightarrow (\Box
\Psi \to P)$. The following tactic performs this step of reasoning.

\begin{minted}{coq}
  destruct FP with (fun A => [! A -> P !]) as (psi, H2).
\end{minted}

\noindent Now, we are able to deduce that $\Psi \to (\Box \Psi\to P)$ from the previous
step result as follows:

\begin{minted}{coq}
  assert (K4; nil |-- [! psi -> []psi -> P !]) as H3.
  apply modal_ax5 in H2; try repeat constructor ;
     try eassumption.
\end{minted}

\noindent The \mintinline{coq}{assert} tactic allows the definition of
auxiliary proof goals. In the previous code piece, we specified
$\vdash \Psi \to (\Box \Psi \to P)$ as a new goal which
generates a new proof hypothesis named \mintinline{coq}{H3}. The
deduction of this fact is immediate from K4 axioms.

@@

\noindent Now, using the axiom \textbf{K} we can deduce $\vdash \Box \Psi \to
\Box (\Box \Psi \to P)$. Again, we deduce this fact using the
\mintinline{coq}{assert} tactic.

\begin{minted}{coq}
  assert (K4; nil |-- [! []psi -> []([]psi -> P) !])
    as H5.
  apply modal_axK in H4; try repeat constructor ;
    try eassumption.
\end{minted}

\noindent The next proof step involves showing that
$\vdash \Box (\Box \Psi \to P)\to (\Box \Box \Psi \to \Box P)$ holds
by axiom \textbf{K}.

\begin{minted}{coq}
  assert (K4; nil |-- [! []([]psi -> P) ->
                      [][]psi -> []P !]) as H6.
  eapply Ax with (a := axK ?[X] ?[Y]);
     try repeat constructor.
\end{minted}

\noindent Using $\vdash \Box (\Box \Psi \to P)\to (\Box \Box \Psi \to \Box P)$ and
$\vdash \Box \Psi \to\Box (\Box \Psi \to P)$ we can deduce
$\vdash \Box \Psi \to \Box \Box \Psi \to \Box P$ using transitivity of
implication.

\begin{minted}{coq}
  assert (K4; nil |-- [! []psi -> [][]psi -> []P !])
    as H7.
  eapply modal_compose; try repeat constructor ;
    try eassumption.
\end{minted}

\noindent Now, we can show that $\vdash\Box\Psi\to\Box\Box\Psi$ using
the axioms of K4. The next snippet of tactics perform this
reasoning step.

\begin{minted}{coq}
  assert (K4; nil |-- [! []psi -> [][]psi !]) as H8.
  apply modal_axK4; try constructor 2.
\end{minted}

\noindent From $\vdash \Box\Psi \to \Box P$ and $\vdash \Box P \to P$, we can
conclude $\vdash \Box \Psi \to P$, by composition of implication as
follows.

\begin{minted}{coq}
  assert (K4; nil |-- [! []psi -> P !]) as H10.
    eapply modal_compose; try repeat constructor ;
      try eassumption.
\end{minted}

\noindent Next, we can conclude $\vdash \Box (\Psi \to P) \to \Psi$ from
$\Psi \leftrightarrow (\Box \Psi \to P)$ and the axioms of K4 using
the following tactics:

\begin{minted}{coq}
  assert (K4; nil |-- [! ([]psi -> P) -> psi !]) as H11.
  apply modal_ax6 in H2; try repeat constructor ;
     try eassumption.
\end{minted}

\noindent From $\vdash (\Box\Psi \to P) \to \Psi$ and $\vdash \Box \Psi \to P$
we can conclude $\vdash \Psi$ by modus ponens.

\begin{minted}{coq}
  assert (K4; nil |-- psi) as H12.
  eapply Mp ; try eassumption.
\end{minted}

\noindent By necessitation rule, we can deduce $\vdash \Box \Psi$

\begin{minted}{coq}
  assert (K4; nil |-- [! []psi !]) as H13.
  apply Nec ; try eassumption.
\end{minted}

\noindent Finally, the desired conclusion $\vdash P$ follows from
using the rule modus ponens over $\vdash \Box \Psi \to P$
and $\vdash \Box \Psi$, as presented by the next tactics
snippet.

\begin{minted}{coq}
  eapply Mp ; try eassumption.
\end{minted}

\noindent This concludes the formalization of the L\"ob's theorem using
the developed library.

@@

A model for recursive, impredicatively quantified types with mutable references
was studied by Appel et. al.~\cite{Appel07}. His interest in formalizing such
advanced typing concepts was in the context of typed assembly languages.
The authors stabilished a semantic type soundness proof by defining a Kripke-style 
semantics of the G\"odel-L\"ob logic of provability. All their definitions were
mechanized through a shallow embedding in Coq.

% The embedding of modal linear logic in Coq was the subject of
% Sadrzadeh's work~\cite{Soton03}. Like our work, Sadrzadeh's uses a
% deep-embedding approach and assumes that all modalities are reflexive
% and transitive. Using her embedding, Sadrzadeh shows some proofs of classical
% logic puzzles.

@@

In this work, we presented a sound deep embedding of arbitrary
normal modal logics in Coq. Our approach is able to represent uniformly
deduction over different modal logic systems by using different axiom sets
as predicate parameter in the deduction inductive type. Using this generic
representation of deduction we are able to prove the soundness theorem and
apply it on proving L\"ob's theorem.
